### 2018-01-29 submission/manuscript.Rmd
nick intro suggestions

Over recent years there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017]. In particular, a lot of the attention of this standardization push has been invested in investigating ways that 16S rRNA gene surveys can be made more replicatable [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016]. Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].  Between the two approaches similar bias considerations include, but are not limited to, preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation. Thus what is identified as a source of bias in 16S rRNA gene sequencing for these overlapping considerations will also influence metagenomic sequencing results.

The most commonly studied biases are by far the skews introduced to the data via preservation media used, storage conditions, and DNA extraction kits chosen. The study of these specific biases has become so large, aggregating them all together has become a difficult task and some researchers provide resources to actively track new findings (e.g. Microbiome Digest - https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/). Within the literature DNA extraction kits have consistently been shown to add bias to downstream analysis [@contamination_Salter2014; @metagenomcis_bias_Costea_2017]. However, the current literature on preservation media and storage conditions has been more mixed, with some studies showing biases while others do not [@preservation_Song_2016; @preservation_Luo_2016; @preservation_Dominianni_2014; @storage_Lauber_2010; @storage_Bassis_2017; @preservation_Sinha_2015]. Although these are important sources of bias they are not the only sources that should be critically examined. The type of DNA polymerase chosen could have a wide ranging affect on downstream results due to error rates and chimeras that may not be easily resolved using bioinformatic approaches.

A recent study in *Nature Biotechnology* showed that there were clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016]. Another important component of this study was that differences, based on DNA polymerase, in the number of OTUs generated were not easily removed using the authors chosen bioinformatic pipeline [@taq_Gohl_2016]. Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerase introduce noticeable biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

We amplified the 16S rRNA gene in both fecal and mock community samples using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA polymerase. First, we tested if we could identify differences in the number of OTUs between the differen HiFi DNA polymerase for both the fecal and mock samples. Next, since there were differences based on HiFi DNA polymerase, we examined if a community measure such as the Bray-Curtis index would be affected. After assessing these community measures we then examined if the different HiFi DNA polymerase had different per base error rates as well as chimera prevalence. Finally, since differences in chimera prevalence, based on HiFi DNA polymerase, could not be completely removed using bioinformatic approaches we tested whether the chimera prevalence correlated well with total number of observed OTUs.

>>>>

Over recent years there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017]. In particular, investigating ways that 16S rRNA gene surveys can be made more replicatable has been a predominant driver of this standardization push [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016]. Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].  Between the two approaches similar bias considerations include, but are not limited to, preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation. Thus, for these overlapping considerations biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.

Currently, preservation media used, storage conditions, and DNA extraction kits chosen are the most commonly studied biases. The study of these specific biases has become so large, aggregating them all together has become a difficult task and some researchers provide resources to actively track new findings (e.g. Microbiome Digest - https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/). Within the literature, DNA extraction kits have consistently been shown to add bias to downstream analysis [@contamination_Salter2014; @metagenomcis_bias_Costea_2017]. However, the current literature on preservation media and storage conditions has been more mixed, with some studies showing biases while others do not [@preservation_Song_2016; @preservation_Luo_2016; @preservation_Dominianni_2014; @storage_Lauber_2010; @storage_Bassis_2017; @preservation_Sinha_2015]. Although these are important sources of bias they are not the only sources that should be critically examined. The type of DNA polymerase chosen could have a wide ranging affect on downstream results due to error rates and chimeras that may not be easily resolved using bioinformatic approaches.

A recent study in *Nature Biotechnology* showed that there were clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016]. Another important component of this study was that differences, based on DNA polymerase, in the number of OTUs generated were not easily removed using the authors chosen bioinformatic pipeline [@taq_Gohl_2016]. Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

We amplified the 16S rRNA gene in both fecal and mock community samples using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA polymerase. First, we tested if we could identify differences in the number of OTUs between the different HiFi DNA polymerase for both the fecal and mock samples. Next, since there were differences based on HiFi DNA polymerase, we examined if a community measure such as the Bray-Curtis index would be affected. After assessing these community measures we then examined if the different HiFi DNA polymerase had different per base error rates as well as chimera prevalence. Finally, since differences in chimera prevalence, based on HiFi DNA polymerase, could not be completely removed using bioinformatic approaches we tested whether the chimera prevalence correlated well with total number of observed OTUs.

### 2018-01-29 submission/manuscript.tex
nick intro suggestions

Over recent years there has been an increasing focus on standardizing
methodological approaches in microbiota research (Kim et al., 2017;
Hugerth \& Andersson, 2017). In particular, a lot of the attention of
this standardization push has been invested in investigating ways that
16S rRNA gene surveys can be made more replicatable (Lauber et al.,
2010; Salter et al., 2014; Song et al., 2016; Gohl et al., 2016).
Although 16S rRNA gene sequencing has been much maligned for introduced
bias, many of these same considerations also affect metagenomic
sequencing (Nayfach \& Pollard, 2016; Costea et al., 2017). Between the
two approaches similar bias considerations include, but are not limited
to, preservation media, storage conditions, DNA extraction kit, PCR, and
sequence library preparation. Thus what is identified as a source of
bias in 16S rRNA gene sequencing for these overlapping considerations
will also influence metagenomic sequencing results.

The most commonly studied biases are by far the skews introduced to the
data via preservation media used, storage conditions, and DNA extraction
kits chosen. The study of these specific biases has become so large,
aggregating them all together has become a difficult task and some
researchers provide resources to actively track new findings
(e.g.~Microbiome Digest -
\url{https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/}).
Within the literature DNA extraction kits have consistently been shown
to add bias to downstream analysis (Salter et al., 2014; Costea et al.,
2017). However, the current literature on preservation media and storage
conditions has been more mixed, with some studies showing biases while
others do not (Lauber et al., 2010; Dominianni et al., 2014; Sinha et
al., 2015; Song et al., 2016; Luo et al., 2016; Bassis et al., 2017).
Although these are important sources of bias they are not the only
sources that should be critically examined. The type of DNA polymerase
chosen could have a wide ranging affect on downstream results due to
error rates and chimeras that may not be easily resolved using
bioinformatic approaches.

A recent study in \emph{Nature Biotechnology} showed that there were
clear differences between normal and high fidelity (HiFi) DNA polymerase
and that you could reduce error and chimera generation by optimizing the
PCR protocol (Gohl et al., 2016). Another important component of this
study was that differences, based on DNA polymerase, in the number of
OTUs generated were not easily removed using the authors chosen
bioinformatic pipeline (Gohl et al., 2016). Although it is probably not
surprising that normal DNA polymerase performed worse than HiFi DNA
polymerase, it is natural to extend this line of inquiry and ask whether
different HiFi DNA polymerase contribute different biases to downstream
sequencing results. There is some reason to think that this may be the
case since many of these HiFi DNA polymerase come from different
families and may intrinsically have different error rates that cannot be
completely removed with modifications (Ishino \& Ishino, 2014). In this
study we critically examine if any of five different HiFi DNA polymerase
introduce noticeable biases into 16S rRNA gene surveys, if this is a
cycle dependent phenomenon, and whether they can be removed using a
standard bioinformatic pipeline.

We amplified the 16S rRNA gene in both fecal and mock community samples
using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA
polymerase. First, we tested if we could identify differences in the
number of OTUs between the differen HiFi DNA polymerase for both the
fecal and mock samples. Next, since there were differences based on HiFi
DNA polymerase, we examined if a community measure such as the
Bray-Curtis index would be affected. After assessing these community
measures we then examined if the different HiFi DNA polymerase had
different per base error rates as well as chimera prevalence. Finally,
since differences in chimera prevalence, based on HiFi DNA polymerase,
could not be completely removed using bioinformatic approaches we tested
whether the chimera prevalence correlated well with total number of
observed OTUs.

>>>>

Over recent years there has been an increasing focus on standardizing
methodological approaches in microbiota research (Kim et al., 2017;
Hugerth \& Andersson, 2017). In particular, investigating ways that 16S
rRNA gene surveys can be made more replicatable has been a predominant
driver of this standardization push (Lauber et al., 2010; Salter et al.,
2014; Song et al., 2016; Gohl et al., 2016). Although 16S rRNA gene
sequencing has been much maligned for introduced bias, many of these
same considerations also affect metagenomic sequencing (Nayfach \&
Pollard, 2016; Costea et al., 2017). Between the two approaches similar
bias considerations include, but are not limited to, preservation media,
storage conditions, DNA extraction kit, PCR, and sequence library
preparation. Thus, for these overlapping considerations biases
identified for 16S rRNA gene sequencing will also likely influence
metagenomic sequencing results.

Currently, preservation media used, storage conditions, and DNA
extraction kits chosen are the most commonly studied biases. The study
of these specific biases has become so large, aggregating them all
together has become a difficult task and some researchers provide
resources to actively track new findings (e.g.~Microbiome Digest -
\url{https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/}).
Within the literature, DNA extraction kits have consistently been shown
to add bias to downstream analysis (Salter et al., 2014; Costea et al.,
2017). However, the current literature on preservation media and storage
conditions has been more mixed, with some studies showing biases while
others do not (Lauber et al., 2010; Dominianni et al., 2014; Sinha et
al., 2015; Song et al., 2016; Luo et al., 2016; Bassis et al., 2017).
Although these are important sources of bias they are not the only
sources that should be critically examined. The type of DNA polymerase
chosen could have a wide ranging affect on downstream results due to
error rates and chimeras that may not be easily resolved using
bioinformatic approaches.

A recent study in \emph{Nature Biotechnology} showed that there were
clear differences between normal and high fidelity (HiFi) DNA polymerase
and that you could reduce error and chimera generation by optimizing the
PCR protocol (Gohl et al., 2016). Another important component of this
study was that differences, based on DNA polymerase, in the number of
OTUs generated were not easily removed using the authors chosen
bioinformatic pipeline (Gohl et al., 2016). Although it is probably not
surprising that normal DNA polymerase performed worse than HiFi DNA
polymerase, it is natural to extend this line of inquiry and ask whether
different HiFi DNA polymerase contribute different biases to downstream
sequencing results. There is some reason to think that this may be the
case since many of these HiFi DNA polymerase come from different
families (e.g. \emph{Taq} belongs to the family A polymerases) and may
intrinsically have different error rates that cannot be completely
removed with modifications (Ishino \& Ishino, 2014). In this study we
critically examine if any of five different HiFi DNA polymerases
introduce significant biases into 16S rRNA gene surveys, if this is a
cycle dependent phenomenon, and whether they can be removed using a
standard bioinformatic pipeline.

We amplified the 16S rRNA gene in both fecal and mock community samples
using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA
polymerase. First, we tested if we could identify differences in the
number of OTUs between the different HiFi DNA polymerase for both the
fecal and mock samples. Next, since there were differences based on HiFi
DNA polymerase, we examined if a community measure such as the
Bray-Curtis index would be affected. After assessing these community
measures we then examined if the different HiFi DNA polymerase had
different per base error rates as well as chimera prevalence. Finally,
since differences in chimera prevalence, based on HiFi DNA polymerase,
could not be completely removed using bioinformatic approaches we tested
whether the chimera prevalence correlated well with total number of
observed OTUs.
